Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2020-12-30 sources/bmc_template_submission/bmc_article.tex
bib file
implementation sections

\parttitle{Conclusions}
The Physcraper workflow demonstrates the benefits of doing open science for
phylogenetics, encouraging researchers to strive for better sharing practices.
Physcraper can be used with any OS and is released under an open-source license.
Detailed instructions for installation and
usage are available at \href{https://physcraper.readthedocs.io}{https://physcraper.readthedocs.io}.

@@

\begin{keyword}
\kwd{gene tree}
\kwd{interoperability}
\kwd{open science}
\kwd{reproducibility}
\kwd{public database}
\kwd{dna alignment}
\end{keyword}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Text and results for this section, as per the individual journal's instructions for authors.

\section*{Implementation}
Text for this section\ldots
\subsection*{Sub-heading for section}
Text for this sub-heading\ldots
\subsubsection*{Sub-sub heading for section}
Text for this sub-sub-heading\ldots
\paragraph*{Sub-sub-sub heading for section}
Text for this sub-sub-sub-heading\ldots

@@

\section*{Figures}
  \begin{figure}[h!]
  \caption{Sample figure title}
\end{figure}

>>>>

\parttitle{Conclusions}
The Physcraper workflow showcases the benefits of doing open science for
phylogenetics, encouraging researchers to strive for better sharing practices.
Physcraper can be used with any OS and is released under an open-source license.
Detailed instructions for installation and
usage are available at \href{https://physcraper.readthedocs.io}{https://physcraper.readthedocs.io}.

@@

\begin{keyword}
\kwd{gene tree}
\kwd{interoperability}
\kwd{open science}
\kwd{reproducibility}
\kwd{public database}
\kwd{DNA alignment}
\end{keyword}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations.
Public biological databases constitute an amazing resource for evolutionary
studies. Updating existing phylogenies with molecular data that has never been
incorporated into any phylogenetic estimate, geographical location, fossils, and
other data in a reproducible and continuous manner is possible by establishing a
data interoperability framework for biological databases. Here, we introduce
Physcraper, a tool that automates database connections to build upon homology
hypotheses that taxon specialists have assessed and deemed appropriate for a
specific phylogenetic scope to update a starting tree and single locus alignments
with public DNA data.

Taxonomic idiosyncrasies across databases represent a huge challenge for automatic
integration of data into phylogenies, which can be addressed with a unified
taxonomy for name standardization. The Open Tree of Life project (OpenTree)
constructs a comprehensive tree of life by synthesizing published phylogenies
and taxonomy. OpenTree's "synthetic" tree comprises 2.3 million tips, of which
around 90,000 are supported by phylogenies - the remaining 1.4 million taxa are
placed in the tree based on taxonomy. To achieve this, OpenTree unifies taxonomic
data from various databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) molecular database GenBank
\cite{benson2000genbank}, among others. The OpenTree taxonomy represents a key resource
for connecting data from any biological database that has been integrated to it.

Another challenge for incorporating public molecular data into existing phylogenies
is assembling high-quality homology hypotheses.
While genomics has, and will continue to, revolutionize phylogenetic inference,
the variety of alternative genomic sequencing approaches it uses produce largely
non-overlapping genomic datasets across taxa, creating challenges in wide scale
phylogenetic reconstruction.

Phylogenomics ameliorate this problem by focusing on targeted capture of informative
loci \cite{andermann2020guide}. Yet, decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that can be used for phylogenetic
reconstruction at many scales.

More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of `r round(236023/240708, digits=4)*100`% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly,
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
However, fine-grained curated markers and alignments can improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

There are almost 8,200 publicly available, peer-reviewed alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be used as seeds to mine molecular databases, and as "jump-start" alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

Physcraper is a Python pipeline using OpenTree's taxonomy and programmatic access
protocols (API's) to implement a database interoperability framework that automatically
links phylogenies that have been standardized to OpenTree taxonomy, to alignments
from TreeBASE, data from GenBank, and phylogenies from OpenTree's Phylesystem.
Physcraper aims to demonstrate the benefits of reproducible workflows and open
science in phylogenetics, and encourage better data sharing practices in the community.

\section*{Implementation}

The general Physcraper framework (Figure 1) consists of 4 steps: 1)identifying and
processing a tree and its underlying alignment; 2) performing a BLAST search of
DNA sequences from original alignment on GenBank, and filtering of new sequences;
3) profile-aligning new sequences to original alignment; 4) performing a phylogenetic
analysis and comparing the updated tree to existing phylogenies.

\subsection*{The inputs: a tree and an alignment}

Taxon names in the input tree must be standardized to OpenTree taxonomy
\cite{rees2017automated} using OpenTree's bulk Taxonomic Name Resolution Service
\href{https://tree.opentreeoflife.org/curator/tnrs/}{TNRS} tool. Users can upload their
own tree, or choose from among the 2, 950 standardized trees stored in OpenTree's
\href{https://github.com/opentreeoflife/phylesystem}{Phylesystem} that also have
alignments available on TreeBASE \cite{piel2009treebase}.

The input alignment is a single locus DNA dataset that was used in part or in
whole to generate the input tree. Physcraper retrieves TreeBASE alignments
automatically. Alternatively, users must provide the path to a local copy of the
alignment.
Only taxa that are both in the sequence alignment and in the tree are considered
further for analysis; at least one taxon and its corresponding sequence are required.

\subsection*{DNA sequence search and filtering}

The Basic Local Alignment Search Tool, BLAST \cite{altschul1990basic} is used for DNA
sequence search on a remote or local GenBank database. It is constrained to a
"search taxon", a taxonomic group in the NCBI taxonomy that is automatically
identified using the OpenTree
\href{<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>}{API}
\cite{rees2017automated}, as the most recent common ancestor of ingroup taxa that is
also a named clade in the NCBI taxonomy (Figure 1).

BLAST is implemented with the `blastn` function \cite{camacho2009blast} and the BioPython
\cite{cock2009biopython} BLAST function from
\href{https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html}{NCBIWWW module}
modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against all DNA sequences in GenBank.
New sequences are excluded for analysis if they 1) are not in the search taxon;
2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a min
and max length threshold, defined as the proportion of the average length without
gaps of all sequences in input alignment (default values of 80% and 120%,
respectively); 4) or if they are either identical to or shorter than an existing
sequence in the input alignment and they represent the same taxon in OpenTree or
NCBI taxonomy.
An arbitrary maximum number of randomly chosen sequences per taxon are allowed
(default to 5).

Reverse, complement, and reverse-complement sequences are identified and translated
using BioPython internal functions \cite{cock2009biopython}.
Iterative cycles of BLAST searches can be performed, by blasting all new sequences
until no new ones are found. By default only one BLAST cycle is performed.

\subsection*{New DNA sequence alignment}

MUSCLE \cite{edgar2004muscle} is used to perform a profile alignment in which the
original alignment is used as a template of homology criteria to align new sequences.
The final alignment is not further automatically checked, and additional inspection
and refinement are recommended.

\subsection*{Tree reconstruction and comparison}

RAxML \cite{stamatakis2014raxml} is implemented to reconstruct a Maximum Likelihood
(ML) gene tree for each input alignment with default settings (GTRCAT model and
100 bootstrap replicates with default algorithm), using input tree as starting
tree for ML searches.
Bootstrap results are summarized using DendroPy's SumTrees module
\cite{sukumaran2010dendropy}.

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
Updated and original tree are compared with Robinson-Foulds weighted and unweighted
metrics estimated with Dendropy \cite{sukumaran2010dendropy}, and with a node by node
comparison between the synthetic OpenTree and original and updated tree individually,
using OpenTree's conflict API \cite{redelings2017supertree}.

@@

\section*{Figures}
  \begin{figure}[h!]
  \caption{The Physcraper framework consists of 4 general steps (see text). The software is fully
  described on its documentation website at \href{https://physcraper.readthedocs.io}{https://physcraper.readthedocs.io},
  along with installation instructions, function usage descriptions, examples and tutorials.
}
\end{figure}
